---
title: "BLUPs for targeted metabolomics - Elite panel"
author: "Malachy Campbell"
date: "8/12/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, root.dir = "/Users/malachycampbell/Documents/Dropbox/Work/Oats/TargetedMet/")
```

## Elite panel

## Preprocessing
First I create two new columns that indicate whether a line is a check or an entry (unreplicated). These are indicated by $check$ and $new$ respectively. The check column consists of c + 1 levels where c is the number of check lines and new consists of two levels binary indicator of whether line is unreplicated entry or check. 

```{r, echo=T, eval = F}
rm(list = ls())

#function for adding check column for each environment. Returns a df of ONLY the env specified
addChks <- function(df = NULL, metaIndx = NULL, ENV = NULL){
  df <- df[df$env %in% ENV ,]
  colIndx <- grep("C\\d{2}.\\d{1}", colnames(df))
  tmp.meta <- df[, -colIndx][metaIndx]
  
  #add columns for GC data
  chks <- names(table(tmp.meta$line)[table(tmp.meta$line) > 2])
  
  tmp.meta <- within(tmp.meta, {
    new <- ifelse(tmp.meta$line %in% chks, 0, 1) #if line is a check then assign 0, else 1
    })
  tmp.meta <- within(tmp.meta, {
    entryc <- ifelse(tmp.meta$new > 0, 999, tmp.meta$line) #checks will have a name, new entries will have 999
    })
  finalDat <- cbind(tmp.meta, df[,colIndx]^(1/3))
  
  return(finalDat)
  }


#Place to store all metabolite data from each stage
FinalData <- list()

FAMEel <- read.csv("Elite/FAME2017_metadata_and_abundance.csv")

FAMEel <- rbind(addChks(df = FAMEel, metaIndx = c(1:9), ENV = "MN"),
                addChks(df = FAMEel, metaIndx = c(1:9), ENV = "WI"),
                addChks(df = FAMEel, metaIndx = c(1:9), ENV = "SD"))

FinalData[["RawData"]] <- FAMEel

saveRDS(FinalData, "Elite/FinalData.rds")
```

Check for outliers
```{r, echo = T, eval = F}
library(sommer)

rm(list = ls())
#data <- readRDS("FinalData.rds")
data <- readRDS("Elite/FinalData.rds")

FAMEel <- data$RawData

FAMEel$block <- as.factor(FAMEel$block)
FAMEel$new <- as.factor(FAMEel$new)
FAMEel$entryc <- as.factor(FAMEel$entryc)
FAMEel$batch <- as.factor(FAMEel$batch)

colIndx <- grep("C\\d{2}.\\d{1}", colnames(FAMEel))

#################
## For GC data ##
#################
ResFAMEel <- list()

dir.create("Elite/OutlierChk")
for(env in c("MN", "SD", "WI")){
  tmpFAMEel <- FAMEel[FAMEel$env %in% env ,]
  
  FAMEelpca <- prcomp(scale(as.matrix(tmpFAMEel[, colIndx]), center = T, scale = T))
  FAMEel.pcs <- cbind(tmpFAMEel[,-colIndx], FAMEelpca$x[,1:9])
  
  tmpRes <- list()
  
  for(cmpd in 1:9){
    tmp.dat <- FAMEel.pcs[c(1:11,(11 + cmpd) )]
    colnames(tmp.dat)[12] <- "Y"
    #tmp.mod <- asreml(fixed = Y ~ entryc, random = ~block + new:line, data = tmp.dat, aom = T)
    full.mod <- mmer(fixed = Y ~ entryc, 
                    random = ~block + new:line + batch, data = tmp.dat)
    stdRes <- resid(full.mod) #standardized residuals
    tmpRes[[cmpd]] <- list(mod = full.mod,
                           stdRes = scale(full.mod$residuals))
  
  pdf(paste0("Elite/OutlierChk/elFAME_PC", env, "_", cmpd, ".pdf"))
  plot(full.mod)
  dev.off()
  
    }
  ResFAMEel[[env]] <- tmpRes
}

saveRDS(ResFAMEel, "Elite/OutlierChk/OutlierCheck.rds")
```

Parese output
```{r, echo = T, eval = F}
rm(list = ls())

Res <- readRDS("Elite/OutlierChk/OutlierCheck.rds")
data <- readRDS("Elite/FinalData.rds")

FAMEel <- data$RawData

findBadSample <- function(results = NULL, thresh = NULL, metadata = NULL, trIndx = NULL, cond = NULL){
  if(cond == "less"){
    indx <- which(results[[trIndx]]$stdRes < thresh)
  }else{
    indx <- which(results[[trIndx]]$stdRes > thresh)
  }
  info <- metadata[indx,]
  return(list(info = info,
              indx = indx))
}

getHat <- function(sommerRes = NULL, trIndx = NULL){
  #taken from sommer plot function to give same results
  #will return leverage given output from sommer mmer function; acts on list so need to provide index
  x <- sommerRes[[trIndx]]$mod
  prov <- mmer(fixed = x$call$fixed, 
               rcov = x$call$rcov, data = x$data, 
        return.param = TRUE, na.method.Y = x$call$na.method.Y, 
        na.method.X = x$call$na.method.X)
  Xlist <- list()
  for (o in 1:length(prov[[3]])) {
        Xlist[[o]] <- kronecker(prov[[2]][[o]], prov[[3]][[o]])
  }
  Xm <- do.call(cbind, Xlist)
  hat <- Xm %*% solve(t(Xm) %*% x$Vi %*% Xm) %*% t(Xm) %*% 
            x$Vi
  return(diag(hat))
}

#MN
findBadSample(metadata = FAMEel[FAMEel$env %in% "MN" ,][,1:11], thresh = -10, 
              trIndx = 3, results = Res$MN, cond = "less")
findBadSample(metadata = FAMEel[FAMEel$env %in% "MN" ,][,1:11], thresh = -4, 
              trIndx = 5, results = Res$MN, cond = "less")
findBadSample(metadata = FAMEel[FAMEel$env %in% "MN" ,][,1:11], thresh = 6, 
              trIndx = 7, results = Res$MN, cond = "greater")
# $info
#                                  pmfname env block column row plot     line CHECK batch new entryc
# 16 16-Metabolomics_17_MN-1-8065-SD120640  MN     1      1   3 8065 SD120640    -1     1   1    999
# 
# $indx
# [1] 16

#WI
findBadSample(metadata = FAMEel[FAMEel$env %in% "WI" ,][,1:11], thresh = -4, 
              trIndx = 1, results = Res$WI, cond = "less")
# $info
#                                    pmfname env block column row plot    line CHECK batch new entryc
# 746 746-Metabolomics_17_WI-12-8285-MN09223  WI    12     15  19 8285 MN09223    -1     5   1    999
# 
# $indx
# [1] 245

MN <- FAMEel[FAMEel$env %in% "MN" ,]; MN <- MN[!row.names(MN) %in% 16 ,]
SD <- FAMEel[FAMEel$env %in% "SD" ,]
WI <- FAMEel[FAMEel$env %in% "WI" ,]; WI <- WI[!row.names(WI) %in% 746 ,]

data[["OutliersRemoved"]] <- rbind(MN, SD, WI)

saveRDS(data, "Elite/FinalData.rds")
```

Calculate BLUPs
```{r, echo = T, eval = F}
library(sommer)

rm(list = ls())

data <- readRDS("Elite/FinalData.rds")

FAMEel <- data$OutliersRemoved

FAMEelBLUPs <- list()
for(env in c("MN", "WI", "SD")){
  tmpFAMEel <- FAMEel[FAMEel$env %in% env ,]
  tmpFAMEel$line <- as.factor(tmpFAMEel$line); tmpFAMEel$line <- droplevels(tmpFAMEel$line)
  tmpFAMEel$block <- as.factor(tmpFAMEel$block); tmpFAMEel$block <- droplevels(tmpFAMEel$block)
  tmpFAMEel$new <- as.factor(tmpFAMEel$new); tmpFAMEel$new <- droplevels(tmpFAMEel$new)
  tmpFAMEel$entryc <- as.factor(tmpFAMEel$entryc); tmpFAMEel$entryc <- droplevels(tmpFAMEel$entryc)
  tmpFAMEel$row <- as.factor(tmpFAMEel$row); tmpFAMEel$row <- droplevels(tmpFAMEel$row)
  tmpFAMEel$column <- as.factor(tmpFAMEel$column); tmpFAMEel$column <- droplevels(tmpFAMEel$column)
  tmpFAMEel$batch <- as.factor(tmpFAMEel$batch); tmpFAMEel$batch <- droplevels(tmpFAMEel$batch)

  colIndx <- grep("C\\d{2}.\\d{1}", colnames(tmpFAMEel))

  #Places to store the results
  BLUPs <- matrix(NA, ncol = length(colIndx), nrow = length(levels(tmpFAMEel$line)))
  drBLUPs <- matrix(NA, ncol = length(colIndx), nrow = length(levels(tmpFAMEel$line)))
  convInfo <- array()
  H2 <- NULL
  
  for(cmpd in 1:length(colIndx)){
    tmp.dat <- tmpFAMEel[c(1:11,(11 + cmpd) )]
    colnames(tmp.dat)[12] <- c("Y")
    
    tmp.mod <- mmer(fixed = Y ~ entryc, 
                    random = ~block + new:line + batch, 
                    data = tmp.dat, na.method.X = "include")
    
    #error handling
    if(length(tmp.mod) < 1){
      H2 <- rbind(H2, data.frame(Estimate = NA, SE = NA))
      BLUPs[,cmpd] <- rep(NA, nrow(BLUPs))
      drBLUPs[,cmpd] <- rep(NA, nrow(BLUPs)) #deregressed blups checks will have NA
      convInfo[cmpd] <- FALSE
    }else{
      PEV <- diag(tmp.mod$PevU$`new:line`$Y)
      PEV <- PEV[grep("new1", names(PEV))]
      varG <- as.numeric(tmp.mod$sigma$`new:line`)
    
      tmp.BLUPs <- tmp.mod$U$`new:line`$Y[grep("new1", names(tmp.mod$U$`new:line`$Y))]
      H2 <- rbind(H2, pin(tmp.mod, h2 ~ V2 / ( V2 + V4)))
      BLUPs[,cmpd] <- tmp.BLUPs #blups
      drBLUPs[,cmpd] <- tmp.BLUPs / (1 - PEV/varG)#deregressed blups checks will have NA
      convInfo[cmpd] <- tmp.mod$convergence
    }
  }
  
  colnames(BLUPs) <- colnames(drBLUPs) <- row.names(H2) <- colnames(tmpFAMEel)[colIndx]
  Lines <- sub("new1:line", "", names(tmp.BLUPs))
  
  FAMEelBLUPs[[env]] <- list(BLUPs = data.frame(line = Lines, BLUPs),
                             drBLUPs = data.frame(line = Lines, drBLUPs),
                             H2 = H2,
                             converged = convInfo)
}

data$BLUPs <- FAMEelBLUPs
saveRDS(data, "Elite/FinalData.rds")
```